Analysis of finite-volume advection schemes on cubed-sphere grids and an accurate \textcolor{red}{scheme} for divergent winds

### 2024-05-20 paper.tex
updated distance notation

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The following geodesic distances in $x$ and $y$ directions, respectively, are introduced:
\begin{align}
	\label{distcube}
	{\delta} x_{ij} = d(\Psi_{i+\frac{1}{2},j},\Psi_{i-\frac{1}{2},j}) , \quad
	{\delta} y_{ij} = d(\Psi_{i,j+\frac{1}{2}},\Psi_{i,j-\frac{1}{2}}),
\end{align}
where $d(P,Q) = R\arccos{(\langle P, Q \rangle)}$, for $P,Q \in \mathbb{S}^2_R$.
These distances may be represented in terms of the tangent vector norms as:
\begin{align}
	{\delta} x_{ij} = 
	\int_{x_{i-\frac{1}{2}}}^{x_{i+\frac{1}{2}}}
	\|\partial_x  {\Psi}_{p}\|(x,y_j) \,dx ,\quad
	{\delta} y_{ij} =
	\int_{y_{j-\frac{1}{2}}}^{y_{j+\frac{1}{2}}}
	\|\partial_y {\Psi}_{p}\|(x_i,y) \,dy.
\end{align}
Hence, their midpoint approximations are defined as:
\begin{align}
	\label{distcube2}
	\hat{\delta} x_{ij} = \|\partial_x {\Psi}_{p}(x_i,y_j)  \|\Delta x,\quad
	\hat{\delta} y_{ij} = \|\partial_y {\Psi}_{p}(x_i,y_j) \|\Delta y.
\end{align}
We point out that $\hat{\delta} x_{ij}$ and $\hat{\delta} y_{ij}$ are replaced in FV3 code by the geodesic distances that they approximate whenever they appear, which are second-order accurate by the midpoint rule.
A control volume of the cubed-sphere is denoted by $\Omega_{ijp}$, defined as $\Omega_{ijp} = {\Psi}_p(\Omega_{ij})$.
The area of $\Omega_{ijp}$ is denoted by $|\Omega_{ij}|$,
since the area does not depend on the panel due to the grid symmetry.
The control volume area may be expressed as:
\begin{equation}
	\label{area}
	|\Omega_{ij}| = \int_{x_{i-\frac{1}{2}}}^{x_{i+\frac{1}{2}}} \int_{y_{j-\frac{1}{2}}}^{y_{j+\frac{1}{2}}}{\sqrt{\mathfrak{g}}(x,y)} \,dx \,dy = 
	|\hat{\Omega}_{ij}| + \mathcal{O}(\Delta x^2),
\end{equation}
where 
\begin{equation}
	\label{area2}
	|\hat{\Omega}_{ij}| = \sqrt{\mathfrak{g}}_{ij} \Delta x \Delta y,
\end{equation} $\sqrt{\mathfrak{g}}_{ij}=\sqrt{\mathfrak{g}}(x_i,y_j) $
and the last equality in Equation \eqref{area} follows from the midpoint rule for integration.
Similar to the grid lengths, the approximated areas $|\hat{\Omega}_{ij}|$ are replaced by the exact area $|\Omega_{ij}|$ in the FV3 code.
\begin{table}[htbp]
	\centering
	\caption{Mean length, minimum length, and maximum length for different values of $N$ considering the equiangular grid.\label{g2-dx-table}}
	\begin{tabular}{|c|c|c|c|c|c}\hline
		$N$ & Mean Length (km) & Min Length (km) & Max Length (km) & $\frac{\text{Max}}{\text{Min}}$ \\ \hline
		48 & 220 & 202 & 240 & 1.1890 \\
		96 & 109 & 99 & 118  & 1.1892 \\
		192 & 54 & 49 & 59  & 1.1892 \\
		384 & 27 & 24 & 29  & 1.1892 \\
		768 & 13 & 12 & 14  & 1.1892\\
		\hline
	\end{tabular} 
\end{table}
\begin{table}[htbp]
	\centering
	\caption{As Table \ref{g2-dx-table} but considering the equi-edge grid. \label{g0-dx-table}}
	\begin{tabular}{|c|c|c|c|c|c}\hline
		$N$ & Mean Length (km) & Min Length (km) & Max Length (km) & $\frac{\text{Max}}{\text{Min}}$ \\ \hline 
		48 & 218 & 175 & 266 & 1.5192 \\
		96 & 108 & 86 & 131 & 1.5195 \\
		192 & 54 & 43 & 65 & 1.5196 \\
		384 & 26 & 21 & 32 & 1.5197 \\
		768 & 13 & 10 & 16 & 1.5197 \\
		\hline
	\end{tabular}
\end{table}

@@

\subsection{The time-averaged and upwind CFL number}
\label{timeav-cfl}
In order to compute the integral in Equation \eqref{numerical-flux}, it will be useful to introduce the time-averaged CFL number at the edges.
This integral in FV3 is expressed in terms of this number.
For the LT2 scheme, the CFL number at the edges is defined naturally as
\begin{align}
	\label{cfl_LT2}
	\tilde{c}_{i+\frac{1}{2}}^{LT2} = {\mathfrak{\tilde{u}}}_{i+\frac{1}{2}}^{LT2}\frac{\Delta t}{\Delta x}
	= {\tilde{u}}_{i+\frac{1}{2}}^{LT2}\frac{\Delta t}{\hat{\delta} x_{i+\frac{1}{2}}},
\end{align}
where ${\tilde{u}}_{i+\frac{1}{2}}^{LT2}$ is just a normalization of 
${\mathfrak{\tilde{u}}}_{i+\frac{1}{2}}^{LT2}$ (Equation \eqref{contra-uv}).
 FV3, on the other hand, uses the upwind CFL approach from \cite{lin:1994}, namely:
\begin{align}
	\label{cfl_fv3}
	{c}_{i+\frac{1}{2}}^{FV3}
    ={\tilde{u}}_{i+\frac{1}{2}}^{FV3}\frac{{\Delta t}}
	{{\hat{\delta} x_{i+\frac{1}{2}}^*}},
\end{align}
where
\begin{align}
	\label{cfl_dp1-dx}
	\hat{\delta} x_{i+\frac{1}{2}}^*  =
	\begin{cases}
		{\hat{\delta} x_{i}},
		\quad &\text{if }{{{u}}_{i+\frac{1}{2}}^{FV3} \ge 0},\\
		{\hat{\delta} x_{i+1}}, 
		\quad &\text{if }{{{u}}_{i+\frac{1}{2}}^{FV3} < 0}.
	\end{cases}
\end{align}
Then Equation \eqref{dp-estimated} may is expressed in terms of the CFL number as:
\begin{equation}
	\label{dp-estimated-cfl}
	\tilde{x}_{i+\frac{1}{2}}^n = x_{i+\frac{1}{2}} - {\tilde{c}}^{M}_{i+\frac{1}{2}}\Delta x,
\end{equation}
where $M=\text{FV3}$ or $M=\text{LT2}$. 
It is easy to see that assuming the absolute value of the CFL number is less than one, then $\tilde{x}_{i+\frac{1}{2}}^n \in [{x}_{i+\frac{1}{2}},{x}_{i+\frac{3}{2}}]$ if ${\tilde{c}}^{M}_{i+\frac{1}{2}}$ is positive, and $\tilde{x}_{i+\frac{1}{2}}^n \in [{x}_{i-\frac{1}{2}},{x}_{i+\frac{1}{2}}]$ otherwise.
Then, as mentioned earlier before, the integral of Equation \eqref{numerical-flux} is constrained to a single control-volume under the CFL condition.
Also, the upwind approach of the numerical flux becomes clear at this point.

@@

\bigg( \frac{\hat{\delta} x_{i+\frac{1}{2}}}{\hat{\delta} x_{i+\frac{1}{2}}^*}\bigg)

@@

		\hat{\delta} y_{i+\frac{1}{2}} 

@@

and recalling the definition of ${\hat{\delta} x_{i+\frac{1}{2}}^*}$ in Equation \eqref{cfl_dp1-dx}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textcolor{red}{
The following geodesic distances in $x$ and $y$ directions, respectively, are introduced:
\begin{align}
	\label{distcube}
	\widehat{d^x_{ij}} = d(\Psi_{i+\frac{1}{2},j},\Psi_{i-\frac{1}{2},j}) , \quad
	\widehat{d^y_{ij}} = d(\Psi_{i,j+\frac{1}{2}},\Psi_{i,j-\frac{1}{2}}),
\end{align}
where $d(P,Q) = R\arccos{(\langle P, Q \rangle)}$, for $P,Q \in \mathbb{S}^2_R$.}
These distances may be represented in terms of the tangent vector norms as:
\begin{align}
	\widehat{d^x_{ij}} = 
	\int_{x_{i-\frac{1}{2}}}^{x_{i+\frac{1}{2}}}
	\|\partial_x  {\Psi}_{p}\|(x,y_j) \,dx ,\quad
	\widehat{d^y_{ij}} =
	\int_{y_{j-\frac{1}{2}}}^{y_{j+\frac{1}{2}}}
	\|\partial_y {\Psi}_{p}\|(x_i,y) \,dy.
\end{align}
Hence, their midpoint approximations are defined as:
\begin{align}
	\label{distcube2}
	{d^x_{ij}} = \|\partial_x {\Psi}_{p}(x_i,y_j)  \|\Delta x,\quad
	{d^y_{ij}} = \|\partial_y {\Psi}_{p}(x_i,y_j) \|\Delta y.
\end{align}
We point out that ${d^x_{ij}}$ and ${d^y_{ij}}$ are replaced in FV3 code by the geodesic distances that they approximate whenever they appear, which are second-order accurate by the midpoint rule.
A control volume of the cubed-sphere is denoted by $\Omega_{ijp}$, defined as $\Omega_{ijp} = {\Psi}_p(\Omega_{ij})$.
The area of $\Omega_{ijp}$ is denoted by $|\Omega_{ij}|$,
since the area does not depend on the panel due to the grid symmetry.
The control volume area may be expressed as:
\begin{equation}
	\label{area}
	|\Omega_{ij}| = \int_{x_{i-\frac{1}{2}}}^{x_{i+\frac{1}{2}}} \int_{y_{j-\frac{1}{2}}}^{y_{j+\frac{1}{2}}}{\sqrt{\mathfrak{g}}(x,y)} \,dx \,dy = 
	|\hat{\Omega}_{ij}| + \mathcal{O}(\Delta x^2),
\end{equation}
where 
\begin{equation}
	\label{area2}
	|\hat{\Omega}_{ij}| = \sqrt{\mathfrak{g}}_{ij} \Delta x \Delta y,
\end{equation} $\sqrt{\mathfrak{g}}_{ij}=\sqrt{\mathfrak{g}}(x_i,y_j) $
and the last equality in Equation \eqref{area} follows from the midpoint rule for integration.
Similar to the grid lengths, the approximated areas $|\hat{\Omega}_{ij}|$ are replaced by the exact area $|\Omega_{ij}|$ in the FV3 code.
\begin{table}[htbp]
	\centering
	\caption{Mean length, minimum length, and maximum length for different values of $N$ considering the equiangular grid.\label{g2-dx-table}}
	\begin{tabular}{|c|c|c|c|c|c}\hline
		$N$ & Mean Length (km) & Min Length (km) & Max Length (km) & $\frac{\text{Max}}{\text{Min}}$ \\ \hline
		48 & 220 & 202 & 240 & 1.1890 \\
		96 & 109 & 99 & 118  & 1.1892 \\
		192 & 54 & 49 & 59  & 1.1892 \\
		384 & 27 & 24 & 29  & 1.1892 \\
		768 & 13 & 12 & 14  & 1.1892\\
		\hline
	\end{tabular} 
\end{table}
\begin{table}[htbp]
	\centering
	\caption{As Table \ref{g2-dx-table} but considering the equi-edge grid. \label{g0-dx-table}}
	\begin{tabular}{|c|c|c|c|c|c}\hline
		$N$ & Mean Length (km) & Min Length (km) & Max Length (km) & $\frac{\text{Max}}{\text{Min}}$ \\ \hline 
		48 & 218 & 175 & 266 & 1.5192 \\
		96 & 108 & 86 & 131 & 1.5195 \\
		192 & 54 & 43 & 65 & 1.5196 \\
		384 & 26 & 21 & 32 & 1.5197 \\
		768 & 13 & 10 & 16 & 1.5197 \\
		\hline
	\end{tabular}
\end{table}

@@

\subsection{The time-averaged and upwind CFL number}
\label{timeav-cfl}
In order to compute the integral in Equation \eqref{numerical-flux}, it will be useful to introduce the time-averaged CFL number at the edges.
This integral in FV3 is expressed in terms of this number.
For the LT2 scheme, the CFL number at the edges is defined naturally as
\begin{align}
	\label{cfl_LT2}
	\tilde{c}_{i+\frac{1}{2}}^{LT2} = {\mathfrak{\tilde{u}}}_{i+\frac{1}{2}}^{LT2}\frac{\Delta t}{\Delta x}
	= {\tilde{u}}_{i+\frac{1}{2}}^{LT2}\frac{\Delta t}
	{{d^x_{i+\frac{1}{2}}}},
\end{align}
where ${\tilde{u}}_{i+\frac{1}{2}}^{LT2}$ is just a normalization of 
${\mathfrak{\tilde{u}}}_{i+\frac{1}{2}}^{LT2}$ (Equation \eqref{contra-uv}).
 FV3, on the other hand, uses the upwind CFL approach from \cite{lin:1994}, namely:
\begin{align}
	\label{cfl_fv3}
	{c}_{i+\frac{1}{2}}^{FV3}
    ={\tilde{u}}_{i+\frac{1}{2}}^{FV3}\frac{{\Delta t}}
	{{{d^{x,*}_{i+\frac{1}{2}}}}},
\end{align}
where
\begin{align}
	\label{cfl_dp1-dx}
    {{d^{x,*}_{i+\frac{1}{2}}}} = 
	\begin{cases}
		{{d^{x}_{i}}},
		\quad &\text{if }{{{u}}_{i+\frac{1}{2}}^{FV3} \ge 0},\\
		{{d^{x}_{i+1}}}, 
		\quad &\text{if }{{{u}}_{i+\frac{1}{2}}^{FV3} < 0}.
	\end{cases}
\end{align}
Then Equation \eqref{dp-estimated} may is expressed in terms of the CFL number as:
\begin{equation}
	\label{dp-estimated-cfl}
	\tilde{x}_{i+\frac{1}{2}}^n = x_{i+\frac{1}{2}} - {\tilde{c}}^{M}_{i+\frac{1}{2}}\Delta x,
\end{equation}
where $M=\text{FV3}$ or $M=\text{LT2}$. 
It is easy to see that assuming the absolute value of the CFL number is less than one, then $\tilde{x}_{i+\frac{1}{2}}^n \in [{x}_{i+\frac{1}{2}},{x}_{i+\frac{3}{2}}]$ if ${\tilde{c}}^{M}_{i+\frac{1}{2}}$ is positive, and $\tilde{x}_{i+\frac{1}{2}}^n \in [{x}_{i-\frac{1}{2}},{x}_{i+\frac{1}{2}}]$ otherwise.
Then, as mentioned earlier before, the integral of Equation \eqref{numerical-flux} is constrained to a single control-volume under the CFL condition.
Also, the upwind approach of the numerical flux becomes clear at this point.

@@

\bigg( \frac{{d}^x_{i+\frac{1}{2}}}{{d}^{x,*}_{i+\frac{1}{2}}}\bigg)

@@

		d^y_{i+\frac{1}{2}} 

@@

and recalling the definition of ${{d}^{x,*}_{i+\frac{1}{2}}}$ in Equation \eqref{cfl_dp1-dx}.
